---
title: "Case Study A"
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
#knitr::opts_knit$set(root.dir = "~/Saez/report")
```

# Introduction

This document is supposed to show how the new implementation can be useful in 
the case of a neighborhood composition analysis.

Loaded packages.

```{r}
library(mistyR, warn.conflicts = FALSE)
library(tidyverse, warn.conflicts = FALSE)
library(future, warn.conflicts = FALSE)
plan("multisession", workers=12)
```

Loading data generated in MIBI_Analysis.Rmd

```{r}
raw.data <- readRDS("data/mibi_tnbc_meta.RDS")
raw.data %>% slice_head(n=6)
```

```{r}
samples <- unique(raw.data$SampleID)
samples
```


Adding a column in Group specifying the immuneGroup

```{r}
meta.data <- raw.data %>%
  select(SampleID, cellSize, Group, immuneGroup, row, col) %>%
  mutate(Group = case_when(
    immuneGroup == "Non-Immune?" ~ Group,
    immuneGroup != "Non-Immune?" ~ immuneGroup,
    TRUE ~ "NA"
  ), immuneGroup = NULL)
meta.data %>% slice_head(n=6)
```

Visualizing the frequency of the cell types.

```{r}
meta.data %>% 
  count(Group) %>%
  ggplot() +
  geom_bar(aes(x=n, y=Group), stat="identity")
```

Prepare the data by one hot encoding the cell types. (removing from each
sample cell types whose total number in the sample in below the `cutoff`)

```{r}
lvls <- meta.data$Group %>% unique
cutoff <- 10

expr.smp <- map(samples, function(smp) {
  raw.expr <- meta.data %>%
    filter(SampleID == smp) %>%
    select(Group) %>%
    mutate(Group = factor(Group, levels = lvls)) %>%
    data.table::as.data.table() %>%
    mltools::one_hot() %>%
    as_tibble() %>%
    select(which(colSums(.) >= cutoff)) %>%
    rename_all(~ str_remove(.x, "Group_")) %>%
    rename_all(make.names)
}) %>% setNames(samples)
expr.smp$`1` %>% slice_head(n=6)
```

Get the positions per sample.

```{r}
positions.smp <- map(samples, function(smp) {
  meta.data %>%
    filter(SampleID == smp) %>%
    select(row, col)
}) %>% setNames(samples)
positions.smp$`1` %>% slice_head(n=6)
```

In how many samples is a given cell types retained?

```{r}
map(expr.smp, ~ colnames(.x)) %>%
  unlist %>%
  tibble() %>%
  rename("celltypes" = ".") %>%
  count(celltypes)
```

Check equivalently in how many cell types are retained in each sample?

```{r fig.height=6}
map2_dfr(expr.smp, names(expr.smp), ~ list("value" = ncol(.x), 
                                           "sample" = .y)) %>%
  ggplot() +
  geom_bar(aes(x = value, y = reorder(sample, value)), stat="identity")
```

Maybe one should remove sample 24.

```{r}
to_remove <- c(24)
#to_remove <- c(5:41) # tmp for testing
expr.smp <- expr.smp[!(names(expr.smp) %in% to_remove)]
positions.smp <- positions.smp[!(names(positions.smp) %in% to_remove)]
```


Computing the MISTy Views, based on the 50 nearest neighbors (constant weight).
(using quietly to capute the message and warnings).

```{r}
generate_misty_views <- function(expr.list, position.list) {
    map2(expr.list, position.list, function(expr, pos) {
    expr %>%
    create_initial_view() %>%
    add_paraview(positions = pos, l = 50, zoi = 0, family = "constant")
  })
}
safe_generate_misty_views <- quietly(generate_misty_views)
```

```{r}
misty.views.run <- safe_generate_misty_views(expr.smp, positions.smp)
misty.views.smp <- misty.views.run$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.views.run$warnings
```
</details>

Running MISTy with the base case.

```{r}
run_misty_wrapper <- function(misty.views.smp) {
  results.paths <- 
  imap(misty.views.smp, function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("output/case_study_A/RF", i), 
              bypass.intra = TRUE,
              cv.folds = 10,
              model.function = random_forest_model)
  })
  collect_results(results.paths)
}
safe_run_misty <- quietly(run_misty_wrapper)
```

```{r}
misty.run.rf <- safe_run_misty(misty.views.smp)
misty.results.rf <- misty.run.rf$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.run.rf$warnings
```
</details>

So essentially I get nothing here, and I think this is due to Ridge regression,
but I am not sure yet.

```{r}
misty.results.rf %>%
  plot_improvement_stats("gain.R2") %>%
  plot_view_contributions() %>%
  plot_interaction_heatmap("para.50")
```

## Regression Boosted Trees

```{r}
run_misty_wrapper <- function(misty.views.smp) {
  results.paths <- 
  imap(misty.views.smp, function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("output/case_study_A/TBOOST", i), 
              bypass.intra = TRUE,
              cv.folds = 10,
               model.function = gradient_boosting_model, booster = "gbtree")
  })
  collect_results(results.paths)
}
safe_run_misty <- quietly(run_misty_wrapper)
```

```{r}
misty.run.tboost <- safe_run_misty(misty.views.smp)
misty.results.tboost <- misty.run.tboost$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.run.tboost$warnings
```
</details>

```{r}
misty.results.tboost %>%
  plot_improvement_stats("gain.R2") %>%
  plot_view_contributions() %>%
  plot_interaction_heatmap("para.50")
```

## Non-negative Least Squares

```{r}
nnls_model <- function(view_data, target, seed, k = 10, ...) {
  
  folds <- withr::with_seed(
    seed,
    caret::createFolds(seq.int(1, nrow(view_data)), k = k)
  )
  
  if (ncol(view_data) <= 3) {
    holdout.predictions <- tibble(index = seq(1, nrow(view_data)),
                                  prediction = view_data %>%
                                    dplyr::pull(tidyselect::all_of(target)) %>%
                                    mean)
    importances <- rep(0, ncol(view_data)-1)
    names(importances) <- colnames(view_data)[colnames(view_data) != target]
  } else {
      holdout.predictions <- purrr::map_dfr(folds, function(holdout) {
    
      in.fold <- seq.int(1, nrow(view_data))[!(seq.int(1, nrow(view_data)) %in%
                                                 holdout)]
      
      train <- view_data[in.fold, ]
      test <- view_data[holdout, ]
      
      pred.train <- train %>% dplyr::select(-tidyselect::all_of(target)) %>%
        as.matrix
      label.train <- train %>% dplyr::pull(tidyselect::all_of(target))
      
      # start changing things here:
      algo.arguments <- list(
        x = pred.train,
        y = label.train,
        alpha = 1,
        lambda = 0,
        lower.limits = 0,
        family = "gaussian"
      )
      
      model <- do.call(glmnet::glmnet, algo.arguments)
      
      pred.test <- test %>% dplyr::select(-tidyselect::all_of(target)) %>% 
        as.matrix
      
      pred <- predict(model, pred.test)
      
      tibble::tibble(index = holdout, prediction = as.vector(pred))
    }) %>% dplyr::arrange(index)
  
    predictors <- view_data %>% 
      dplyr::select(-tidyselect::all_of(target)) %>% as.matrix
    labels <- view_data %>% dplyr::pull(tidyselect::all_of(target))
    
       # start changing things here:
      algo.arguments.wm <- list(
        x = predictors,
        y = labels,
        alpha = 1,
        lambda = 0,
        lower.limits = 0,
        family = "gaussian"
      )
    
    whole.model <- do.call(glmnet::glmnet, algo.arguments.wm)
    
    whole.model$beta
    
    importances <- whole.model$beta[,1]
    # fix for bypass intra (replace NA with 0 for consistent behavior)
    importances <- ifelse(is.na(importances), 0, importances)
  }
  
  list(unbiased.predictions = holdout.predictions, 
       importances = importances)
}
```

```{r}
run_misty_wrapper <- function(misty.views.smp) {
  results.paths <- 
  imap(misty.views.smp, function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("output/case_study_A/NNLS", i), 
              bypass.intra = TRUE,
              cv.folds = 10,
               model.function = nnls_model)
  })
  collect_results(results.paths)
}
safe_run_misty <- quietly(run_misty_wrapper)
```

```{r}
misty.run.nnls <- safe_run_misty(misty.views.smp)
misty.results.nnls <- misty.run.nnls$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.run.nnls$warnings
```
</details>

```{r}
misty.results.nnls %>%
  plot_improvement_stats("gain.R2") %>%
  plot_view_contributions() %>%
  plot_interaction_heatmap("para.50")
```

## Classification Random Forest

```{r}
random_forest_class <- function(view_data, target, seed, ...) {
  
  ellipsis.args <- list(...)

  # If we have no predictors simply return the mean of the target
  if (ncol(view_data) <= 2) {
    predictions <- tibble(index = seq(1, nrow(view_data)),
                          prediction = view_data %>%
                            dplyr::pull(tidyselect::all_of(target)) %>%
                                    mean)
    importances <- rep(0, ncol(view_data)-1)
    names(importances) <- colnames(view_data)[colnames(view_data) != target]
  } else {
  
    target.vector <- view_data %>% dplyr::pull(target) %>% as.factor
  
    transformed_view_data <- view_data %>%
      mutate(!!target := target.vector)
    
    # default ranger arguments
    algo.arguments <- list(
      formula = stats::as.formula(paste0(target, " ~ .")),
      data = transformed_view_data,
      num.trees = 100,
      importance = "impurity",
      mtry = NULL,
      verbose = FALSE,
      num.threads = 1,
      seed = seed
    )
  
    if (!(length(ellipsis.args) == 0)) {
      algo.arguments <- merge_two(algo.arguments, ellipsis.args)
    }
  
    model <- do.call(ranger::ranger, algo.arguments)
  
    predictions <- tibble::tibble(
      index = seq_len(nrow(view_data)),
      prediction = as.numeric(levels(model$predictions))[model$predictions]
    )
    importances = model$variable.importance
  }
  list(
    unbiased.predictions = predictions,
    importances = importances
  )
}
```

```{r}
run_misty_wrapper <- function(misty.views.smp) {
  results.paths <- 
  imap(misty.views.smp, function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("output/case_study_A/RFCLASS", i), 
              bypass.intra = TRUE,
              cv.folds = 10,
               model.function = random_forest_class)
  })
  collect_results(results.paths)
}
safe_run_misty <- quietly(run_misty_wrapper)
```

```{r}
misty.run.rfc <- safe_run_misty(misty.views.smp)
misty.results.nnls <- misty.run.rfc$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.run.rfc$warnings
```
</details>

```{r}
misty.results.nnls %>%
  plot_improvement_stats("gain.R2") %>%
  plot_view_contributions() %>%
  plot_interaction_heatmap("para.50")
```

# Appendix

```{r eval=FALSE, include=FALSE}
map_lgl(samples[!(samples %in% to_remove)], function(smp) {
  #print(paste0("output/case_study_A/RFCLASS", smp, "/performance.txt"))
  read.table(paste0("output/case_study_A/RFCLASS", smp, "/performance.txt"))
  TRUE
})

map_lgl(samples[!(samples %in% to_remove)], function(smp) {
  #print(paste0("output/case_study_A/RFCLASS", smp, "/coefficients.txt"))
  read.table(paste0("output/case_study_A/RFCLASS", smp, "/coefficients.txt"))
  TRUE
})
```


```{r eval=FALSE, include=FALSE}
views <- misty.views.smp$`1`
target <- "B"
target.vector <- expr %>% dplyr::pull(target)
model.function <- nnls_model
model.name <- "nnls"
cv.folds <- 10
bypass.intra <- TRUE
seed <- 42
cached <- FALSE
ellipsis.args <- list()
ellipsis.args.text <- ""
view <- views$intraview
transformed.view.data <-
            tibble::tibble(!!target := target.vector, ".novar" := 0)
view_data <- transformed.view.data
```

```{r eval=FALSE, include=FALSE}
bypass.intra <- TRUE
model.function <- random_forest_model
cached <- FALSE
views <- misty.views.smp$`1`
results.folder <- "results"
seed <- 42
target.subset <- NULL
append <- FALSE
ellipsis.args <- list()
model.name <- "rf"
cv.folds <- 10
target <- "CD8.T"
target <- "Keratin.positive.tumor"
```

```{r eval=FALSE, include=FALSE}
misty.views <- misty.views.smp$`1`
expr <- misty.views$intraview$data
target = "Neutrophils"
target.vector <- expr %>% dplyr::pull(target)
transformed.view.data1 <-
            tibble::tibble(!!target := target.vector, ".novar" := 0)

transformed.view.data2 <- misty.views$paraview.50$data %>%
  mutate(!!target := target.vector)

view_data <- transformed.view.data2
```

```{r eval=FALSE, include=FALSE}
views <- misty.views
target <- "Neutrophils"
model.function <- random_forest_class
model.name <- "rf_class"
cv.folds <- 10
bypass.intra <- TRUE
seed <- 42
cached <- FALSE
view <- views$intraview
ellipsis.args.text <- ""
colnames(view$data)
target <- "Keratin.positive.tumor"
```





