---
title: "Case Study A"
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

# Introduction

This document is supposed to show how the new implementation can be useful in the case of a neighborhood composition analysis. Meaning whether we can predict the identiy of a cell type using the neighboring cell types.

However, after updating the MISTy meta model from linear regression to ridge regression, MISTy is not able to learn something from the spatial context anymore.

# Setup

## Packages

Loaded packages.

```{r}
library(mistyR, warn.conflicts = FALSE)
library(tidyverse, warn.conflicts = FALSE)
library(future, warn.conflicts = FALSE)
plan("multisession", workers=12)
```

## Import

Loading data generated in "eda_mibi_tnbc.Rmd".

```{r}
raw.data <- readRDS("data/mibi_tnbc_meta.RDS")
raw.data %>% slice_head(n=6)
```

How many samples do we have?

```{r}
samples <- unique(raw.data$SampleID)
samples
```

Adding a column in Group specifying the immuneGroup.

```{r}
meta.data <- raw.data %>%
  select(SampleID, cellSize, Group, immuneGroup, row, col) %>%
  mutate(Group = case_when(
    immuneGroup == "Non-Immune?" ~ Group,
    immuneGroup != "Non-Immune?" ~ immuneGroup,
    TRUE ~ "NA"
  ), immuneGroup = NULL)
meta.data %>% slice_head(n=6)
```

Simplifying the cell types.

```{r}
meta.data <- meta.data %>%
  mutate(Group = case_when(
    Group == "DC/Mono" | Group == "DC" ~ "DC",
    Group == "Keratin-positive tumor" | Group == "Tumor" ~ "Tumor",
    TRUE ~ Group
  ))
```

Visualizing the frequency of the cell types.

```{r}
meta.data %>% 
  count(Group) %>%
  ggplot() +
  geom_bar(aes(x=n, y=Group), stat="identity") +
  labs(x = "n", y = "Celltype")
```

We see already that tumor cells make up by far the most cell types. Such unbalanced datasets pose problems for many machine learning algorithms, as the prediction error can easily be reduced by always predicting the most abundant class. 

We will remove NK cells because they are so rare, and Unidentified cells / Other immune because they are not very helpful.

```{r}
meta.data <- meta.data %>%
  filter(!(Group %in% c("Unidentified", "NK", "Other immune")))
  #filter(!(Group %in% c("Unidentified", "NK", "Other immune", "Tumor")))
```

Prepare the data by one hot encoding the cell types. Also we will remove from each sample cell types whose total number in the sample in below the `cutoff`, which is here `10`. 

```{r}
lvls <- meta.data$Group %>% unique
cutoff <- 10

expr.smp <- map(samples, function(smp) {
  raw.expr <- meta.data %>%
    filter(SampleID == smp) %>%
    select(Group) %>%
    mutate(Group = factor(Group, levels = lvls)) %>%
    data.table::as.data.table() %>%
    mltools::one_hot() %>%
    as_tibble() %>%
    select(which(colSums(.) >= cutoff)) %>%
    rename_all(~ str_remove(.x, "Group_")) %>%
    rename_all(make.names)
}) %>% setNames(samples)
expr.smp$`1` %>% slice_head(n=6)
```

Get the positions per sample.

```{r}
positions.smp <- map(samples, function(smp) {
  meta.data %>%
    filter(SampleID == smp) %>%
    select(row, col)
}) %>% setNames(samples)
positions.smp$`1` %>% slice_head(n=6)
```

In how many samples is a given cell types retained?

```{r}
map(expr.smp, ~ colnames(.x)) %>%
  unlist %>%
  tibble() %>%
  rename("celltypes" = ".") %>%
  count(celltypes)
```

We see that especially DC (dendritic cells), B cells, NK (natural killer cells), and Tregs (regulatory T-cells) are only present in a quarter or half of the samples.

Check equivalently how many cell types are retained in each sample?

```{r fig.height=6}
map2_dfr(expr.smp, names(expr.smp), ~ list("value" = ncol(.x), 
                                           "sample" = .y)) %>%
  ggplot() +
  geom_bar(aes(x = value, y = reorder(sample, value)), stat="identity")
```

We will remove sample 24, since it only contrain 4 distinct cell types.

```{r}
to_remove <- c(24)
#to_remove <- c(5:41) # tmp for testing
expr.smp <- expr.smp[!(names(expr.smp) %in% to_remove)]
positions.smp <- positions.smp[!(names(positions.smp) %in% to_remove)]
```

## MISTy Views

Computing the MISTy Views, based on the 50 nearest neighbors (constant weight).
(using quietly to capute the message and warnings).

```{r}
generate_misty_views <- function(expr.list, position.list) {
    map2(expr.list, position.list, function(expr, pos) {
    expr %>%
    create_initial_view() %>%
    add_paraview(positions = pos, l = 50, zoi = 0, family = "constant")
  })
}
safe_generate_misty_views <- quietly(generate_misty_views)

misty.views.run <- safe_generate_misty_views(expr.smp, positions.smp)
misty.views.smp <- misty.views.run$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.views.run$warnings
```
</details>

# Running MISTy

## Random Forest

Running MISTy with the base case.

```{r include=FALSE}
misty.views.smp$`1`$paraview.50$data[1,] %>% unlist %>% plot
```


```{r}
run_misty_wrapper <- function(misty.views.smp) {
  results.paths <- 
  imap(misty.views.smp, function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("output/case_study_A/RF", i), 
              bypass.intra = TRUE,
              cv.folds = 10,
              model.function = random_forest_model)
  })
  collect_results(results.paths)
}
safe_run_misty <- quietly(run_misty_wrapper)

misty.run.rf <- safe_run_misty(misty.views.smp)
misty.results.rf <- misty.run.rf$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.run.rf$warnings
```
</details>

As mentioned in the introduction, MISTy is not able to learn anything from the spatial context.

```{r}
misty.results.rf %>%
  plot_improvement_stats("gain.R2") %>%
  plot_view_contributions() %>%
  plot_interaction_heatmap("para.50")
```

## Regression Boosted Trees

```{r}
run_misty_wrapper <- function(misty.views.smp) {
  results.paths <- 
  imap(misty.views.smp, function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("output/case_study_A/TBOOST", i), 
              bypass.intra = TRUE,
              cv.folds = 10,
               model.function = gradient_boosting_model, booster = "gbtree")
  })
  collect_results(results.paths)
}
safe_run_misty <- quietly(run_misty_wrapper)

misty.run.tboost <- safe_run_misty(misty.views.smp)
misty.results.tboost <- misty.run.tboost$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.run.tboost$warnings
```
</details>



```{r}
misty.results.tboost %>%
  plot_improvement_stats("gain.R2") %>%
  plot_view_contributions() %>%
  plot_interaction_heatmap("para.50")
```

## Non-negative Least Squares

```{r}
nnls_model <- function(view_data, target, seed, k = 10, ...) {
  
  folds <- withr::with_seed(
    seed,
    caret::createFolds(seq.int(1, nrow(view_data)), k = k)
  )
  
  if (ncol(view_data) <= 3) {
    holdout.predictions <- tibble(index = seq(1, nrow(view_data)),
                                  prediction = view_data %>%
                                    dplyr::pull(tidyselect::all_of(target)) %>%
                                    mean)
    importances <- rep(0, ncol(view_data)-1)
    names(importances) <- colnames(view_data)[colnames(view_data) != target]
  } else {
      holdout.predictions <- purrr::map_dfr(folds, function(holdout) {
    
      in.fold <- seq.int(1, nrow(view_data))[!(seq.int(1, nrow(view_data)) %in%
                                                 holdout)]
      
      train <- view_data[in.fold, ]
      test <- view_data[holdout, ]
      
      pred.train <- train %>% dplyr::select(-tidyselect::all_of(target)) %>%
        as.matrix
      label.train <- train %>% dplyr::pull(tidyselect::all_of(target))
      
      # start changing things here:
      algo.arguments <- list(
        x = pred.train,
        y = label.train,
        alpha = 1,
        lambda = 0,
        lower.limits = 0,
        family = "gaussian"
      )
      
      model <- do.call(glmnet::glmnet, algo.arguments)
      
      pred.test <- test %>% dplyr::select(-tidyselect::all_of(target)) %>% 
        as.matrix
      
      pred <- predict(model, pred.test)
      
      tibble::tibble(index = holdout, prediction = as.vector(pred))
    }) %>% dplyr::arrange(index)
  
    predictors <- view_data %>% 
      dplyr::select(-tidyselect::all_of(target)) %>% as.matrix
    labels <- view_data %>% dplyr::pull(tidyselect::all_of(target))
    
       # start changing things here:
      algo.arguments.wm <- list(
        x = predictors,
        y = labels,
        alpha = 1,
        lambda = 0,
        lower.limits = 0,
        family = "gaussian"
      )
    
    whole.model <- do.call(glmnet::glmnet, algo.arguments.wm)
    
    whole.model$beta
    
    importances <- whole.model$beta[,1]
    # fix for bypass intra (replace NA with 0 for consistent behavior)
    importances <- ifelse(is.na(importances), 0, importances)
  }
  
  list(unbiased.predictions = holdout.predictions, 
       importances = importances)
}
```

```{r}
run_misty_wrapper <- function(misty.views.smp) {
  results.paths <- 
  imap(misty.views.smp, function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("output/case_study_A/NNLS", i), 
              bypass.intra = TRUE,
              cv.folds = 10,
               model.function = nnls_model)
  })
  collect_results(results.paths)
}
safe_run_misty <- quietly(run_misty_wrapper)

misty.run.nnls <- safe_run_misty(misty.views.smp)
misty.results.nnls <- misty.run.nnls$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.run.nnls$warnings
```
</details>

Just like for the random forest model, there is no gain in R2 here.

```{r}
misty.results.nnls %>%
  plot_improvement_stats("gain.R2") %>%
  plot_view_contributions() %>%
  plot_interaction_heatmap("para.50")
```

## Classification Random Forest

```{r}
random_forest_class <- function(view_data, target, seed, ...) {
  
  ellipsis.args <- list(...)

  # If we have no predictors simply return the mean of the target
  if (ncol(view_data) <= 2) {
    predictions <- tibble(index = seq(1, nrow(view_data)),
                          prediction = view_data %>%
                            dplyr::pull(tidyselect::all_of(target)) %>%
                                    mean)
    importances <- rep(0, ncol(view_data)-1)
    names(importances) <- colnames(view_data)[colnames(view_data) != target]
  } else {
  
    target.vector <- view_data %>% dplyr::pull(target) %>% as.factor
  
    transformed_view_data <- view_data %>%
      mutate(!!target := target.vector)
    
    # default ranger arguments
    algo.arguments <- list(
      formula = stats::as.formula(paste0(target, " ~ .")),
      data = transformed_view_data,
      num.trees = 100,
      importance = "impurity",
      mtry = NULL,
      verbose = FALSE,
      num.threads = 1,
      seed = seed
    )
  
    if (!(length(ellipsis.args) == 0)) {
      algo.arguments <- merge_two(algo.arguments, ellipsis.args)
    }
  
    model <- do.call(ranger::ranger, algo.arguments)
  
    predictions <- tibble::tibble(
      index = seq_len(nrow(view_data)),
      prediction = as.numeric(levels(model$predictions))[model$predictions]
    )
    importances = model$variable.importance
  }
  list(
    unbiased.predictions = predictions,
    importances = importances
  )
}
```

```{r}
run_misty_wrapper <- function(misty.views.smp) {
  results.paths <- 
  imap(misty.views.smp, function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("output/case_study_A/RFCLASS", i), 
              bypass.intra = TRUE,
              cv.folds = 10,
               model.function = random_forest_class)
  })
  collect_results(results.paths)
}
safe_run_misty <- quietly(run_misty_wrapper)

misty.run.rfc <- safe_run_misty(misty.views.smp)
misty.results.nnls <- misty.run.rfc$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.run.rfc$warnings
```
</details>

```{r}
misty.results.nnls %>%
  plot_improvement_stats("gain.R2") %>%
  plot_view_contributions() %>%
  plot_interaction_heatmap("para.50")
```

# Appendix

```{r eval=FALSE, include=FALSE}
map_lgl(samples[!(samples %in% to_remove)], function(smp) {
  #print(paste0("output/case_study_A/RFCLASS", smp, "/performance.txt"))
  read.table(paste0("output/case_study_A/RFCLASS", smp, "/performance.txt"))
  TRUE
})

map_lgl(samples[!(samples %in% to_remove)], function(smp) {
  #print(paste0("output/case_study_A/RFCLASS", smp, "/coefficients.txt"))
  read.table(paste0("output/case_study_A/RFCLASS", smp, "/coefficients.txt"))
  TRUE
})
```

```{r eval=FALSE, include=FALSE}
views <- misty.views.smp$`1`
target <- "B"
target.vector <- expr %>% dplyr::pull(target)
model.function <- nnls_model
model.name <- "nnls"
cv.folds <- 10
bypass.intra <- TRUE
seed <- 42
cached <- FALSE
ellipsis.args <- list()
ellipsis.args.text <- ""
view <- views$intraview
transformed.view.data <-
            tibble::tibble(!!target := target.vector, ".novar" := 0)
view_data <- transformed.view.data
```

```{r eval=FALSE, include=FALSE}
bypass.intra <- TRUE
model.function <- random_forest_model
cached <- FALSE
views <- misty.views.smp$`1`
results.folder <- "results"
seed <- 42
target.subset <- NULL
append <- FALSE
ellipsis.args <- list()
model.name <- "rf"
cv.folds <- 10
target <- "CD8.T"
target <- "Keratin.positive.tumor"
```

```{r eval=FALSE, include=FALSE}
misty.views <- misty.views.smp$`1`
expr <- misty.views$intraview$data
target = "Neutrophils"
target.vector <- expr %>% dplyr::pull(target)
transformed.view.data1 <-
            tibble::tibble(!!target := target.vector, ".novar" := 0)

transformed.view.data2 <- misty.views$paraview.50$data %>%
  mutate(!!target := target.vector)

view_data <- transformed.view.data2
```

```{r eval=FALSE, include=FALSE}
views <- misty.views
target <- "Neutrophils"
model.function <- random_forest_class
model.name <- "rf_class"
cv.folds <- 10
bypass.intra <- TRUE
seed <- 42
cached <- FALSE
view <- views$intraview
ellipsis.args.text <- ""
colnames(view$data)
target <- "Keratin.positive.tumor"
```





