---
title: "Case Study B"
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE)
knitr::opts_knit$set(root.dir = "~/Saez/report")
```

# Introduction

This document is supposed to show how the new implementation can be useful in 
the case of a neighborhood composition analysis.

Loaded packages.

```{r}
library(mistyR, warn.conflicts = FALSE)
library(tidyverse, warn.conflicts = FALSE)
library(future, warn.conflicts = FALSE)
plan("multisession", workers=12)
```

```{r}
raw.expr <- readRDS("data/expression.RDS")
raw.data <- readRDS("data/mibi_tnbc_meta.RDS")
meta.data <- raw.data %>%
  select(SampleID, cellSize, Group, immuneGroup, row, col) %>%
  mutate(Group = case_when(
    immuneGroup == "Non-Immune?" ~ Group,
    immuneGroup != "Non-Immune?" ~ immuneGroup,
    TRUE ~ "NA"
  ), immuneGroup = NULL)
meta.data %>% slice_head(n=6)
```

```{r}
samples <- unique(meta.data$SampleID)
samples
```


Getting expression per sample.

```{r}
expr.smp <- map(samples, function(smp) {
    ret.expr <- raw.expr[meta.data$SampleID == smp, ] %>%
      # Select only proteins
      select(11:47)
    
    # Check for zero variance (otherwise MISTy throws an error)
    target.var <- apply(ret.expr, 2, stats::sd, na.rm = TRUE)
    ret.expr <- ret.expr %>% select(-names(which(target.var == 0)))
    
    # Check for how many unique values
    target.unique <- colnames(ret.expr) %>%
      purrr::set_names() %>%
      purrr::map_int(~ length(unique(ret.expr %>% pull(.x))))
    
    colnames(ret.expr) <- make.names(colnames(ret.expr))
    ret.expr
  }) %>% setNames(samples)
expr.smp$`1` %>% slice_head(n=6)
```

Getting positions for each sample

```{r}
pos.smp <- map(samples, function(smp) {
  meta.data %>%
    filter(SampleID == smp) %>%
    select(row, col)
}) %>% set_names(samples)
pos.smp$`1` %>% slice_head(n=6)
```

Compute MISTy views.

```{r message=FALSE, warning=FALSE}
misty.views.smp <- map2(expr.smp, pos.smp, function(expr, coord) {
  create_initial_view(expr) %>%
    add_juxtaview(positions = coord, neighbor.thr = 40) %>%
    add_paraview(positions = coord, l = 120, zoi = 40)
}) %>% set_names(samples)
```

Run classic model.

```{r}
run_misty_wrapper <- function(misty.views.smp) {
  results.paths <- 
  imap(misty.views.smp, function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("output/case_study_B/RF", i), 
              bypass.intra = TRUE,
              cv.folds = 10,
              model.function = random_forest_model)
  })
  collect_results(results.paths)
}
safe_run_misty <- quietly(run_misty_wrapper)
```

```{r}
misty.run.rf <- safe_run_misty(misty.views.smp)
rf.results <- misty.run.rf$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.run.rf$warnings
```
</details>

Look at the results.

```{r}
rf.results %>%
  plot_improvement_stats("multi.R2") %>%
  plot_improvement_stats("gain.R2") %>%
  plot_view_contributions() %>%
  plot_interaction_heatmap("juxta.40", clean=TRUE, cutoff=1.5) %>%
  plot_interaction_heatmap("para.120", clean=TRUE)
```

Non-negative least squares for modeling the views.

```{r}
nnls_model <- function(view_data, target, seed, k = 10, ...) {
  
  folds <- withr::with_seed(
    seed,
    caret::createFolds(seq.int(1, nrow(view_data)), k = k)
  )
  
  check_unique <- all(map_lgl(folds, function(holdout) {
    in.fold <- seq.int(1, nrow(view_data))[!(seq.int(1, nrow(view_data)) %in%
                                                 holdout)]
    train <- view_data[in.fold, ]
    all(map_dbl(train, ~ length(unique(.x))) > 1)
  }))
  
  if (ncol(view_data) <= 3 | !check_unique) {
    holdout.predictions <- tibble(index = seq(1, nrow(view_data)),
                                  prediction = view_data %>%
                                    dplyr::pull(tidyselect::all_of(target)) %>%
                                    mean)
    importances <- rep(0, ncol(view_data)-1)
    names(importances) <- colnames(view_data)[colnames(view_data) != target]
  } else {
      holdout.predictions <- purrr::imap_dfr(folds, function(holdout, i) {
    
      in.fold <- seq.int(1, nrow(view_data))[!(seq.int(1, nrow(view_data)) %in%
                                                 holdout)]
      
      train <- view_data[in.fold, ]
      test <- view_data[holdout, ]
      
      pred.train <- train %>% dplyr::select(-tidyselect::all_of(target)) %>%
        as.matrix
      label.train <- train %>% dplyr::pull(tidyselect::all_of(target))
      
      # start changing things here:
      algo.arguments <- list(
        x = pred.train,
        y = label.train,
        alpha = 1,
        lambda = 0,
        lower.limits = 0,
        family = "gaussian"
      )
      
      model <- do.call(glmnet::glmnet, algo.arguments)
      
      pred.test <- test %>% dplyr::select(-tidyselect::all_of(target)) %>% 
        as.matrix
      
      pred <- predict(model, pred.test)
      
      tibble::tibble(index = holdout, prediction = as.vector(pred))
    }) %>% dplyr::arrange(index)
  
    predictors <- view_data %>% 
      dplyr::select(-tidyselect::all_of(target)) %>% as.matrix
    labels <- view_data %>% dplyr::pull(tidyselect::all_of(target))
    
       # start changing things here:
      algo.arguments.wm <- list(
        x = predictors,
        y = labels,
        alpha = 1,
        lambda = 0,
        lower.limits = 0,
        family = "gaussian"
      )
    
    whole.model <- do.call(glmnet::glmnet, algo.arguments.wm)
    
    importances <- whole.model$beta[,1]
    # fix for bypass intra (replace NA with 0 for consistent behavior)
    importances <- ifelse(is.na(importances), 0, importances)
  }
  
  list(unbiased.predictions = holdout.predictions, 
       importances = importances)
}
```

Run nnls model.

```{r}
run_misty_wrapper <- function(misty.views.smp) {
  results.paths <- 
  imap(misty.views.smp, function(misty.views, i) {
  
  misty.views %>%
    run_misty(results.folder = paste0("output/case_study_B/NNLS", i), 
              bypass.intra = TRUE,
              cv.folds = 10,
              model.function = nnls_model)
  })
  collect_results(results.paths)
}
safe_run_misty <- quietly(run_misty_wrapper)
```

```{r}
misty.run.nnls <- safe_run_misty(misty.views.smp)
nnls.results <- misty.run.rf$result
```

<details> 
<summary>Click to see warnings</summary>
```{r}
misty.run.rf$warnings
```
</details>

```{r}
nnls.results %>%
  plot_improvement_stats("multi.R2") %>%
  plot_improvement_stats("gain.R2") %>%
  plot_view_contributions() %>%
  plot_interaction_heatmap("juxta.40", clean=TRUE, cutoff = 1.5) %>%
  plot_interaction_heatmap("para.120", clean=TRUE)
```

# Appendix

```{r eval=FALSE, include=FALSE}
# tmp
target <- "Lag3"
results.folder <- paste0("output/case_study_B/RF", 1)
views <- misty.views.smp$`2`
model.function <- random_forest_model
model.name <- "rf"
cv.folds <- 10
bypass.intra <- FALSE
seed <- 42
cached <- FALSE
append <- FALSE
target.subset <- c()
ellipsis.args <- list()
ellipsis.args.text <- ""
views$intraview$data %>% colnames
target <- "OX40"
```

```{r eval=FALSE, include=FALSE}
# tmp
results.folder <- paste0("output/case_study_B/NNLS", 2)
views <- misty.views.smp$`2`
model.function <- nnls_model
model.name <- "nnls"
cv.folds <- 10
bypass.intra <- FALSE
seed <- 42
cached <- FALSE
append <- FALSE
target.subset <- c()
ellipsis.args <- list()
ellipsis.args.text <- ""
views$intraview$data %>% colnames
target <- "OX40"
view <- views$intraview

view_data <- transformed.view.data
target <- "OX40"
seed <- 42
k <- 10
```

