---
title: "EDA MIBI TNBC"
output:
  workflowr::wflow_html:
    toc: yes
    toc_float: yes
    toc_depth: 2
    code_folding: hide
editor_options: 
  chunk_output_type: inline
bibliography: references.bib
csl: science.csl
---

Setup.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE)
#knitr::opts_knit$set(root.dir = "~/Saez/report")
```

```{r}
suppressPackageStartupMessages(library(mistyR))
suppressPackageStartupMessages(library(tidyverse))
```

# Loading and Processing Data

The columns Group and ImmuneGroup in the cell metadata are translated from
numbers to their actual meaning.

```{r}
input <- read_csv("data/mibi_tnbc/cellData.csv", show_col_types=FALSE)

# See description in Readme
group.trans <- c("1" = "Unidentified", "2" = "Immune", "3" = "Endothelial", 
                 "4" = "Mesenchymal-like", "5" = "Tumor", 
                 "6" = "Keratin-positive tumor")

immune.group.trans <- c("0" = "Non-Immune?", "1" = "Tregs", "2" = "CD4 T", "3" = "CD8 T", 
                        "4" = "CD3 T", "5" = "NK", "6" = "B",
                        "7" = "Neutrophils", "8" = "Macrophages", 
                        "9" = "DC", "10" = "DC/Mono", "11" = "Mono/Neu", 
                        "12" = "Other immune")

raw_data <- input %>%
  mutate(Group = group.trans[as.character(Group)]) %>%
  mutate(immuneGroup = immune.group.trans[as.character(immuneGroup)])
  
raw_data %>% slice_head(n=6)
```

Next we add the coordinates of the cells.

```{r}
if ("coord.RDS" %in% list.files("data/mibi_tnbc")) {
  coord <- readRDS("data/mibi_tnbc/coord.RDS")
} else {
  library(tiff)
  lbs <- c(1:29, 31:41)
  coord <- furrr::future_map_dfr(lbs, function(id) {
    print(id)
    # Read in raw matrix (in tiff format)
    tiff <- readTIFF(paste0("/home/philipp/data/saez/mibi/p", id, "_labeledcellData.tiff"),
                     as.is = TRUE)
    seq.rows <- seq_len(nrow(tiff))
    seq.cols <- seq_len(ncol(tiff))
    # important: map over all unique values here! (but I removed 1 and 2)
    cell.ids <- unique(as.vector(tiff))[-which(unique(as.vector(tiff)) %in% c(0,1))]
    map_dfr(cell.ids, function(i) {
      if (i %% 100 == 0) print(i)
      # Convert to binary matrix with TRUE and FALSE
      binary <- (tiff == i)
      s <- sum(binary)
      # Calculate center of mass
      c(id = id, 
        i = i,
        x.center = sum(seq.rows * rowSums(binary)) / s,
        y.center = sum(seq.cols * colSums(binary)) / s
      )
    })
  })
  saveRDS(coord, file = "data/mibi_tnbc/coord.RDS")
}

# join coordinates with the meta data.
data <- raw_data %>%
  inner_join(coord, by = c("SampleID" = "id", 
                                       "cellLabelInImage" = "i")) %>%
  rename(row = x.center, col = y.center)
```

However, we have some coordinates for cell IDs which are not in the meta data.
For sample 40 there are for example 572 of those cell IDs which can be 
found in the tiff but not in the meta data. Maybe those spots turned out not 
be cells in the end (segmentation errors?). Or these are the cells with low 
expression across all markers ("sum(expression) < 0.1").

```{r}
dplyr::setdiff(
  coord %>% filter(id == 40) %>% pull(i),
  data %>% filter(SampleID == 40) %>% pull(cellLabelInImage)
  ) %>%
  length()
```

```{r echo=FALSE}
# Before proceeding with the analysis, check whether we have any NAs.
stopifnot(sum(map_int(colnames(data), ~ sum(is.na(data[[.x]])))) == 0)
```

Split the data into expression matrix and meta data.

```{r}
meta <- data %>%
  select(c(1:3, 53:59))
meta %>% slice_head(n=6)

expr <- data %>%
  select(4:52)
expr %>% slice_head(n=6)
```

# Meta Data

## Cell Counts

Number of cells per sample. Curiously sample 30 is missing.

```{r}
meta %>%
  count(SampleID) %>%
  ggplot() +
  geom_bar(aes(x=SampleID, y=n), stat = "identity")
```

Number of "cell types" over all samples as annotated by the authors or the
study. We mostly have tumor and immune cells. Keratin serves as a marker
for cancer cells.

```{r}
meta %>%
  count(Group) %>%
  arrange(desc(n))
```

Next we will have a closer look at the immune populations. And according
to the number of cells present, macrophages and T-cells are the major
players, followed by B-cells.

```{r}
meta %>%
  count(immuneGroup) %>%
  arrange(desc(n))
```

## Patient/Sample Metadata

Patient data provided with the raw data.

```{r}
patient.info <- read_csv("data/mibi_tnbc/patient_class.csv",
                         col_names = c("SampleID", "Classification"),
                         show_col_types=FALSE)
patient.info %>% slice_head(n=6)
```

```{r}
patient.info.paper <- readxl::read_xlsx("data/mibi_tnbc/mmc2.xlsx", 
                                        skip = 1, col_names = TRUE)

patient.info.paper %>% slice_head(n=6)
```

# Count Data

Histograms per marker over all samples. We can see that that there is a huge
variability in the expression of different markers. The measured entities 
other than proteins (Au, Background, Beta catentin, C, Ca, Fe, Si, Ta) were
later excluded from the analysis with MISTy.

How were the "counts" normalized?

According to the readme.rtf the markers were normalized by dividing the 
counts by the cell size and transforming with arcsinh (inverse hyperbolic sine).
Furthermore in the paper it is stated that after arcsinh transformation the
markers were standard normalized.

```{r fig.height=12, fig.width=18}
expr %>%
  pivot_longer(cols = everything(), names_to = "marker", values_to = "expression") %>%
  ggplot() +
  geom_histogram(aes(x=expression), bins = 100) +
  facet_wrap(~ marker, scales = "free")
```

Looking at the boxplot for the expression per marker over all samples, ignoring
outliers which are defined as points which includes all points outside of 
median ± 1.5 * IQR (interquantile range), we can again see the broad
variability.

```{r}
expr %>%
  pivot_longer(cols = everything(), names_to = "marker", values_to = "expression") %>%
  ggplot() +
  geom_boxplot(aes(y=expression, x=marker), outlier.shape = NA) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  coord_cartesian(ylim = c(-5, 5))
```

Focusing on the measured proteins, the most highly expressed species 
(according to the mean) are:

```{r}
tibble(marker = colnames(expr), mean.expression = colMeans(expr)) %>%
  filter(!(marker %in% c("Ta", "Au", "P", "Si", "C", "Na", "Ca", "Fe", 
                         "dsDNA", "Backgrond"))) %>%
  slice_max(mean.expression, n=6)
```

# Spatial Data

## General

What is the scale? - 1 Pixel corresponds to about 390 nm.

We will first look at the spatial data from all samples in one plot. The
cell types are annotated according to the information the authors of the 
original study provided.

As pointed out in their publication, one can see that there is a continuum 
between mixing of immune cells (yellow) and cancer tissue (green and blue).
Either we can see a clear interface between both populations or they are
totally mixed. This is actually associated with survival (see paper).
Below this observation is quantified by calculating a similar mixing score
as they did in the original publication.

```{r fig.height=12, fig.width=18}
meta %>%
  ggplot() +
  geom_point(aes(x = row, y = col, col = Group), size = .1) +
  facet_wrap(~ SampleID) + 
  guides(colour = guide_legend(override.aes = list(size = 4)))
```

## Mixing Score

Calculating the mixing score by counting for all immune cells in each
sample how many contacts to they have to cancer cells divided by the 
contacts to other immune cells. The contacts are determined by
Delaunay Triangulation and cells are only considered to be neighbors
if the distance between them is below 40. 

Importantly this mixing score is higher if there are more tumor cells
than immune cells (as seen by the filling color of the barplot below,
`ratio.count.tumor.immune` refers to the number of tumor cells 
divided by the number of immune cells).

(Remark: The authors of the paper used Haralick’s gray-level co-occurrence
matrix to determine direct neighbors of each cell)

```{r}
# Setting the distance threshold for what is considered a neighbor in the DT.
neighbor.thr <- 40

if ("tumor.immune.contacts.RDS" %in% list.files("data/mibi_tnbc")) {
  contacts.smp <- readRDS("data/mibi_tnbc/tumor.immune.contacts.RDS")
} else {
  
  coord.type.smp <- unique(meta$SampleID) %>%
  set_names(paste0("s", .)) %>%
  map(function(id) {
    meta %>%
      filter(meta$SampleID == id) %>% 
      select(row, col, Group) %>%
      mutate(Group = ifelse(Group == "Keratin-positive tumor", 
                            "Tumor", Group)) %>%
      filter(Group %in% c("Tumor", "Immune")) %>%
      mutate(id = seq_len(nrow(.)))
  })

  # helper function for getting direct neighbors
  get_neighbors <- function(ddobj, id) {
    dplyr::union(
      ddobj$delsgs$ind1[which(ddobj$delsgs$ind2 == id)],
      ddobj$delsgs$ind2[which(ddobj$delsgs$ind1 == id)]
    )
  }

  contacts.smp <- coord.type.smp %>%
  set_names() %>%
  map_dfr(function(smp) {
    
    delaunay <- deldir::deldir(as.data.frame(smp %>% select(row, col)))
    dists <- distances::distances(as.data.frame(smp %>% select(row, col)))
    ids.to.map <- smp %>% filter(Group == "Immune") %>% pull(id)
    
    map(ids.to.map, function(i) {
      alln <- get_neighbors(delaunay, i)
      actualn <- alln[which(dists[alln, i] <= neighbor.thr)]
      
      smp %>%
        filter(id %in% actualn) %>%
        pull(Group)
      
    }) %>% 
      unlist %>%
      table
  })
  saveRDS(contacts.smp, "data/mibi_tnbc/tumor.immune.contacts.RDS")
}

meta %>%
  group_by(SampleID) %>%
  count(Group) %>% 
  pivot_wider(names_from = Group, values_from=n, names_prefix = "Count.") %>%
  cbind(contacts.smp) %>%
  mutate(ratio.tumor.immune = as.vector(Tumor/Immune * 100)) %>%
  rename("Count.Keratin.positive.Tumor" = "Count.Keratin-positive tumor") %>%
  replace(is.na(.), 0) %>%
  mutate(Count.Tumor.Total = Count.Keratin.positive.Tumor + Count.Tumor) %>%
  mutate(ratio.count.tumor.immune = Count.Tumor.Total / Count.Immune) %>%
  ggplot() +
  geom_bar(aes(x=reorder(SampleID, -ratio.tumor.immune), 
               y=ratio.tumor.immune,
               fill=ratio.count.tumor.immune), 
           stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Sample", y = "Mixing Score") +
  scale_fill_viridis_c()
```

Looking at the samples with the top mixing score, namely 15, 18, 22, 24, 31, 
25, 21, 26

```{r fig.width=12, fig.height=6}
meta %>%
  filter(SampleID %in% c(15, 18, 22, 24, 31, 25, 21, 26)) %>%
  ggplot() +
  geom_point(aes(x = row, y = col, col = Group), size = .1) +
  facet_wrap(~ SampleID, ncol = 4) + 
  guides(colour = guide_legend(override.aes = list(size = 4)))
```

Looking at the samples with the lowest mixing score, namely 35, 28, 16, 37, 
4, 40, 41, 5.

```{r fig.width=12, fig.height=6}
meta %>%
  filter(SampleID %in% c(35, 28, 16, 37, 4, 40, 41, 5)) %>%
  ggplot() +
  geom_point(aes(x = row, y = col, col = Group), size = .1) +
  facet_wrap(~ SampleID, ncol = 4) + 
  guides(colour = guide_legend(override.aes = list(size = 4)))
```

And we can see that those mixing scores resemble exactly what we see
in the images.

## Sample 40

Looking more closely at sample 40 to make sure that the preprocessing
went right. First, we will look at the cells only using their
calculated center of mass.

```{r fig.height=6, fig.width=8}
meta %>%
  filter(SampleID == 40) %>%
  ggplot() +
  geom_point(aes(x = row, y = col, col = Group), size=.4)  + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
```

Furthermore, we will look at the tiff.

```{r fig.height=6, fig.width=8}
library(tiff)
id = 40
tiff <- readTIFF(paste0("data/mibi_tnbc/raw_tiffs/p", 
                        id, "_labeledcellData.tiff"), as.is = TRUE)
```

First, the binary tiff, meaning all values that there not 0 or 1 (i.e. a
cell was detected).

```{r fig.height=6, fig.width=7}
# get coordinates for where there was a cell detected (=: binary)
binary <- apply(tiff, c(1,2), function(n) !(n %in% c(0,1)))
binary.df <- as.data.frame(binary)
rownames(binary.df) <- seq_len(nrow(binary))
colnames(binary.df) <- seq_len(ncol(binary))

df <- binary.df %>% 
  rownames_to_column(var = "row") %>%
  pivot_longer(cols = !(row), names_to = "col") %>%
  mutate(row = as.numeric(row), col = as.numeric(col))

subs <- sample(nrow(df), .03 * nrow(df))

df %>%
  slice(n = subs) %>%
  mutate(cell.present = value) %>%
  ggplot() +
  geom_point(aes(x=row, y=col, col=cell.present), size=.1) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()) +
  scale_colour_manual(values = c("TRUE" = "green2", 
                                 "FALSE" = "blue4"))  + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
```

Second, we will visualize the tiff and the labeled center of mass together.
And on this scale the result actually looks quite nice.

```{r fig.height=6, fig.width=8}
s.40 <- meta %>%
  filter(SampleID == 40)

tiff.40 <- df %>%
  filter(value) %>%
  slice_sample(., n = nrow(.)/10)

ggplot() +
  geom_point(data = tiff.40, aes(x=row, y=col), col="black", 
             size = 0.1, stroke = 0, shape = 16) +
  geom_point(data = s.40, aes(x=row, y=col, col=Group), size = .5) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank())  + 
  guides(colour = guide_legend(override.aes = list(size = 2)))
```

To qualitatively assess the performance of the segmentation we will furthermore
look at a close up of sample 40, more specifically the region from row 0 to
row 600 and from col 1448 to col 2048. And we can see
that the center of mass (colored point) mostly matches the cell boundaries,
but not always!

```{r fig.height=8, fig.width=10}
c.max <- dim(binary)[1]
top_left <- list(row = 0, col = c.max)
bottom_right <- list(row = 600, col = c.max-600)

# get the unique coordinates for each cell
tiff.df <- as.data.frame(tiff)
rownames(tiff.df) <- seq_len(nrow(tiff))
colnames(tiff.df) <- seq_len(ncol(tiff))

tiff.df.plot <- tiff.df %>%
  rownames_to_column(var = "row") %>%
  pivot_longer(cols = !(row), names_to = "col") %>%
  mutate(row = as.numeric(row), col = as.numeric(col)) %>%
  filter(!(value %in% c(0, 1))) %>%
  # how to get only entry for each unique value
  group_by(value) %>%
  filter(row_number()==1) %>%
  ungroup()

s.40 <- meta %>%
  filter(SampleID == 40) %>%
  filter(row >= top_left$row, row <= bottom_right$row,
         col >= bottom_right$col, col <= top_left$col)

tiff.40 <- df %>%
  filter(value) %>%
  filter(row >= top_left$row, row <= bottom_right$row,
         col >= bottom_right$col, col <= top_left$col)

labels.40 <- tiff.df.plot %>%
  filter(row >= top_left$row, row <= bottom_right$row,
         col >= bottom_right$col, col <= top_left$col)

ggplot() +
  geom_point(data = tiff.40, aes(x=row, y=col), col="black", 
             size = 0.1, stroke = 0, shape = 16) +
  geom_point(data = s.40, aes(x=row, y=col, col=Group), size = 3) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank())
```

Here are the corresponding IDs for comparison to the meta data.

```{r fig.height=7, fig.width=8}
ggplot(labels.40) +
  geom_label(aes(x = row, y = col, label = value), size = 2.5)
```

# Merging Metadata

Merging the meta data with the mixing score and some other measures.

```{r}
# Generating Sample Meta Data
immune.counts <- meta %>%
  group_by(SampleID) %>%
  count(Group) %>%
  pivot_wider(names_from = Group, values_from = n,
              names_prefix = "Count.")

total.counts <- meta %>%
  count(SampleID) %>%
  rename(Count.Total = n)

mixing.score <- round(
  contacts.smp %>% 
    mutate(ratio.tumor.immune = Tumor/Immune * 100) %>%
    pull(ratio.tumor.immune),
  2)

sample.meta <- patient.info %>%
  filter(SampleID != 30) %>%
  mutate(ratio.tumor.immune = as.vector(mixing.score)) %>%
  inner_join(immune.counts, by = "SampleID") %>%
  inner_join(total.counts, by = "SampleID") %>%
  # see paper Figure 3, Panel C
  mutate(Classification = 
         case_when(Classification == 0 ~ "mixed",
                   Classification == 1 ~ "compartimentalized",
                   Classification == 2 ~ "cold")) %>% 
  mutate(SampleID = as.character(SampleID)) %>%
  left_join(patient.info.paper, by = c("SampleID" = "InternalId"))

sample.meta
```

Helper function to calculate distances to the neighbors.

```{r}
# helper function for nearest neighbors
dist_k_nn <- function(sample, k) {
  nn <- RANN::nn2(sample, eps=0.0, k = k)
  map_dbl(seq_len(nrow(sample)), function(n) {
    mean(nn$nn.dists[n, ])
  })
}

# helper function for getting direct neighbors
get_neighbors <- function(ddobj, id) {
  dplyr::union(
    ddobj$delsgs$ind1[which(ddobj$delsgs$ind2 == id)],
    ddobj$delsgs$ind2[which(ddobj$delsgs$ind1 == id)]
  )
}
```

First we will look at the distribution of the  mean distances to the direct 
neighbors (as determined by the delauney triangulation) to get a feeling for 
which cutoff value should be used for the generation of the juxtaview. The
corresponding distribution is shown below.
(in blue the cutoff - 40 pixels (15.6 microns) - which was used in the end).

```{r}
coord.smp <- map(unique(coord$id), function(smp){
  coord %>%
    filter(id == smp) %>%
    select(x.center, y.center)
})

# mean distance to the direct neighbors for each cell
mean.dists.direct.neighbors <- map(coord.smp, function(smp) {
  delaunay <- deldir::deldir(as.data.frame(smp))
  dists <- distances::distances(as.data.frame(smp))
  map_dbl(seq_len(nrow(smp)), function(id) {
    nn <- get_neighbors(delaunay, id)
    mean(dists[nn, id])
  })
}) 

stat.per.sample <- map2_dfr(mean.dists.direct.neighbors,
                            seq_len(length(mean.dists.direct.neighbors)),
                            function(list, smp) {
            data.frame(sample = smp, value = list)
          })

# histogram
ggplot(data.frame(dist = mean.dists.direct.neighbors %>% unlist)) +
  geom_histogram(aes(dist), bins = 100) +
  lims(x = c(0,100)) +
  geom_vline(xintercept = 40, col = "blue")
```

Summary statistics:

```{r}
data.frame(value = mean.dists.direct.neighbors %>% unlist) %>%
  summarise(mean = mean(value), 
            median = median(value), 
            q25 = quantile(value, 0.25),
            q75 = quantile(value, 0.75),
            min = min(value),
            max = max(value))
```

However, a more detailed look across the sample (boxplot) shows that the
cutoff may be too low for certain samples. It may be good to increase this
threshold to 50.

```{r}
# boxplot
stat.per.sample %>%
  mutate(sample = factor(sample)) %>%
  ggplot() +
  geom_boxplot(aes(x=sample, y=value)) +
  geom_hline(yintercept = 40, col="blue") +
  lims(y = c(0, 200))
```

Furthermore we will look at the distribution of the mean distances to the
10 -, 20 -, 40 -, and 80 closest cells to get a feeling
for which cutoff will be used to generate the paraview 
(in blue the cutoff - 120 pixels (46.8 microns) - which was used in the end).

```{r}
neighbor.dists <- map_dfr(coord.smp, function(smp) {
  c(10, 20, 40, 80) %>%
    purrr::set_names() %>%
    map_dfc(function(k) {
  dist_k_nn(smp, k)
    }) %>%
  purrr::set_names(paste0("dist_to_", c("10", "20", "40", "80")))
})

neighbor.dists %>%
  pivot_longer(cols = everything()) %>%
  mutate(name = factor(name, 
                       levels = paste0("dist_to_", c("10", "20", "40", "80")))) %>%
  ggplot() +
  geom_histogram(aes(value), bins = 100) +
  facet_wrap(~ name) +
  lims(x = c(0, 200)) +
  geom_vline(xintercept = 120, col = "blue")
```

Summary statistics:

```{r}
neighbor.dists %>%
  pivot_longer(cols = everything()) %>%
  group_by(name) %>%
  summarise(mean = mean(value), 
            median = median(value), 
            q25 = quantile(value, 0.25),
            q75 = quantile(value, 0.75),
            min = min(value),
            max = max(value))
```

