---
title: "EDA MERFISH LIVER"
output:
  workflowr::wflow_html:
    toc: yes
    toc_float: yes
    toc_depth: 2
    code_folding: hide
editor_options: 
  chunk_output_type: inline
bibliography: references.bib
csl: science.csl
---

Setup.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE)
knitr::opts_knit$set(root.dir = "~/Saez/report")
```

Loaded Packages.

```{r}
suppressPackageStartupMessages(library(mistyR))
suppressPackageStartupMessages(library(future))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(R.matlab))
plan("multisession")
```

# Introduction

The data originate from the following study: 

Lu, Y. et al. Spatial transcriptome profiling by MERFISH reveals fetal liver 
hematopoietic stem cell niche architecture. Cell Discov 7, 47 (2021).

# WT Data

## Gene Information

```{r}
gene.tibble <- readxl::read_xlsx(
  "data/merfish_liver/Supplement/41421_2021_266_MOESM1_ESM.xlsx", 
  skip = 1
  )
genes <- gene.tibble$`Gene name`
gene.tibble %>% slice_head(n=6)
```

## Cell Data

Raw Data:

```{r eval = FALSE}
files <- list.files("data/merfish_liver/Data/WT niche/", full.names = TRUE)

sample.names <- str_extract(files, "(?<=//).+(?=.mat)")
raw.data <- map(files, function(file) {
  readMat(file)$CellList
  }) %>% setNames(sample.names)
```


```{r}
wt.samples <- list.files("data/merfish_liver/Data/WT CellList")
sample.names <- str_extract(wt.samples, "CellID_[0-9]+")

if ("merfish.raw.data.RDS" %in% list.files("data/merfish_liver")) {
  raw.data <- readRDS("data/merfish_liver/merfish.raw.data.RDS")
} else {

  raw.data <- map(wt.samples, function(file) {
  readMat(paste0("data/merfish_liver/Data/WT CellList/",
                             file))$CellList
  }) %>% setNames(sample.names)
  
  saveRDS(raw.data, "data/merfish_liver/merfish.raw.data.RDS")
}
```

```{r}
sample.names
```

Cell Metadata:

```{r}
cell.data <- map(raw.data, function(smp.data) {
  smp.cell.data <- map_dfr(seq.int(1, dim(smp.data)[3]), function(i) {
    cell_i <- smp.data[1:8, 1, i]
    tibble::tibble(id = cell_i$CellID, fov = cell_i$FOV, x = cell_i$Center[1, 1],
                   y = cell_i$Center[1, 2], total.rna.copy = cell_i$TotalRNACopyNumber[1, 1], 
           edge = cell_i$OnEdge[1, 1], type = cell_i$CellType, num.pixel = dim(cell_i$PixelList)[1])
  })
}) %>% setNames(sample.names)
```

```{r}
map(cell.data, ~ nrow(.x))

#cell.data$CellID_190411 %>% distinct(fov, .keep_all= TRUE)

#cell.data$CellID_190411 %>% group_by(fov) %>% count
```

Merging Cell Data

```{r}
cell.meta <- cell.data %>% map2_dfr(names(cell.data), function(smp.data, name) {
  smp.data %>% mutate(sample = name)
}) %>%
  mutate(type_name = case_when(
  type == 1 ~ "Arterial.Endothelial.Cells",
  type == 2 ~ "Sinusoidal.Endothelial.Cells", 
  type == 3 ~ "Megakaryocytes",
  type == 4 ~ "Hepatocyte",
  type == 5 ~ "Macrophage",
  type == 6 ~ "Myeloid",
  type == 7 ~ "Erythroid.Progenitor",
  type == 8 ~ "Erythroid.Cell",
  type == 9 ~ "Unknown"
))
cell.meta %>% slice_head(n=6)
```

Expression Data:

```{r}
expr.data <- map(raw.data, function(smp.data) {
  smp.expr.data <- map_dfr(seq.int(1, dim(smp.data)[3]), function(i) {
    cell_i <- smp.data[1:8, 1, i]
    cell_i$CellID
    expr <- cell_i$RNACopyNumber %>% as.vector
    names(expr) <- genes
    c(id = cell_i$CellID, expr)
  })
})
```

```{r}
expr.data$CellID_190411 %>% slice_head(n=6)
```

Normalizing the expression data.

We divide each count by the sum of counts in the cell and then multiply by 1000
(just as they did it in the paper).

```{r}
expr.data.norm <- map(expr.data, function(sample) {
  cell.totals <- sample %>%
    pivot_longer(cols = 2:141) %>%
    group_by(id) %>%
    summarise(cell.sum = sum(value))
  
  sample %>%
    left_join(cell.totals, by="id") %>%
    pivot_longer(cols = 2:141) %>%
    mutate(normalized = value / cell.sum * 1000) %>%
    select(-c(cell.sum, value)) %>%
    pivot_wider(names_from = name, values_from = normalized)
})
```

```{r}
expr.data.norm$CellID_190411 %>% slice_head(n = 6)
```

Looking at the distribution for the top 36 markers by sum of expression. And
even for these markers we mostly have zero counts.

```{r fig.width=14, fig.height=10}
expr.data$CellID_190411 %>%
  select(- id) %>%
  pivot_longer(cols = everything()) %>%
  group_by(name) %>%
  mutate(sum.marker = sum(value)) %>%
  ungroup() %>%
  slice_max(order_by = sum.marker, n = 23700*36) %>%
  ggplot() +
  geom_histogram(aes(value), bins = 30) +
  facet_wrap(~ name, ncol=6, nrow=6, scales="free")
```

Cell Types per Sample

```{r}
cell.meta %>%
  count(sample, type_name) %>%
  ggplot() +
  geom_bar(aes(y = type_name, x = n), stat="identity") +
  facet_wrap(~ sample) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Visualization of a Part of the Sample

The visualization are now based on the assumption that the images are
1536 x 1536 (actually there is no proper documentation)

```{r}
if ("CellID_190411.pixels.RDS" %in% list.files("data/merfish_liver")) {
  CellID_190411.pixels <- readRDS("data/merfish_liver/CellID_190411.pixels.RDS")
} else {
  img.size <- 1536
  CellID_190411.pixels <- map_dfr(seq.int(1, dim(raw.data$CellID_190411)[3]), function(i) {
    cell_i <- raw.data$CellID_190411[1:8, 1, i]
    tibble::tibble(id = cell_i$CellID, type = cell_i$CellType, fov = cell_i$FOV, 
                   pixels = cell_i$PixelList)
  }) %>%
  mutate(type = factor(type)) %>%
  mutate(id = factor(id)) %>%
  mutate(y = pixels %% img.size, x = ceiling(pixels / img.size))
  
  saveRDS(CellID_190411.pixels, "data/merfish_liver/CellID_190411.pixels.RDS")
}
rm(raw.data)
```

Verification of the right image size.

Example field of view 5.

```{r}
CellID_190411.pixels %>%
  filter(fov == 5) %>%
  ggplot() +
  geom_point(aes(x=x, y=y, col=id), size=.1) +
  theme(legend.position = "none") +
  scale_color_manual(values=rep(c("dodgerblue", "brown", "purple", "blue",
                                  "seagreen", "gold", "tomato", "turquoise",
                                  "magenta", "springgreen"), 250)) +
  coord_equal()
```

Example field of view 1.

```{r}
CellID_190411.pixels %>%
  filter(fov == 1) %>%
  ggplot() +
  geom_point(aes(x=x, y=y, col=id), size=.1) +
  theme(legend.position = "none") +
  scale_color_manual(values=rep(c("dodgerblue", "brown", "purple", "blue",
                                  "seagreen", "gold", "tomato", "turquoise",
                                  "magenta", "springgreen"), 250)) +
  coord_equal()
```

Check wether the center of mass makes sense. It looks very good.

```{r}
center.mass <- cell.data$CellID_190411 %>%
  filter(fov == 5) %>%
  mutate(id = factor(id)) %>%
  mutate(center_x = x, center_y = y) %>%
  select(c(id, fov, center_x, center_y))

CellID_190411.pixels %>%
  filter(fov == 5) %>%
  left_join(center.mass, by="id") %>%
  ggplot() +
  geom_point(aes(x=x, y=y, col=id), size=.1, alpha=0.01) +
  geom_point(aes(x=center_x, y=center_y, col=id), size=2) +
  coord_equal()
```

Since once cannot really say how and whether it makes sense to stitch those
images together, we will run MISTy on each sample on each field of view.

But first we will remove all the FOVs which have too few cells.

```{r}
map2_dfr(cell.data, names(cell.data), function(smp, name) {
  smp %>%
    group_by(fov) %>%
    summarise(n = n()) %>%
    mutate(sample = name)
}) %>%
  ggplot() +
  geom_histogram(aes(n), bins=50, col="black", fill="blue") +
  facet_wrap(~ sample) +
  geom_vline(xintercept = 125, col="red")
```

Removing every FOV with fewer than 280 cells

```{r}
thr <- 280
keep.smp <- map(cell.data, function(smp) {
  smp %>%
  group_by(fov) %>%
  summarise(n = n()) %>%
  filter(n > thr) %>%
  pull(fov) %>%
  as.numeric()
})

map_dbl(keep.smp, ~ length(.x))
```

Let's look at the average distance to the direct neighbors per cell per field of view
(per sample).

```{r}
# helper function for nearest neighbors
dist_k_nn <- function(sample, k) {
  nn <- RANN::nn2(sample, eps=0.0, k = k)
  map_dbl(seq_len(nrow(sample)), function(n) {
    mean(nn$nn.dists[n, ])
  })
}

# helper function for getting direct neighbors
get_neighbors <- function(ddobj, id) {
  dplyr::union(
    ddobj$delsgs$ind1[which(ddobj$delsgs$ind2 == id)],
    ddobj$delsgs$ind2[which(ddobj$delsgs$ind1 == id)]
  )
}
```

So the best threshold for the juxtaview may be 100.

```{r}
# map over FOV in each sample
mean.dists.direct.neighbors.smp <- map2(keep.smp, names(keep.smp), function(keep, name) {
  map(keep, function(test.fov) {
    coords <- cell.data[[name]] %>% filter(fov == test.fov) %>%
    select(c(x, y)) %>% as.data.frame
    delaunay <- deldir::deldir(coords)
    dists <- distances::distances(coords)
    map_dbl(seq_len(nrow(coords)), function(id) {
      nn <- get_neighbors(delaunay, id)
      mean(dists[nn, id])
    })
  })
})

map2_dfr(mean.dists.direct.neighbors.smp, names(mean.dists.direct.neighbors.smp), function(dists, name) {
  tibble::tibble(dist = unlist(dists), sample = name)
}) %>%
  ggplot() +
  geom_histogram(aes(dist), bins = 100) +
  lims(x = c(0,200)) +
  geom_vline(xintercept = 100, col = "blue") +
  facet_wrap(~ sample)
```

Furthermore we will look at the distribution of the mean distances to the 
10 -, 20 -, 30 -, and 40 closest cells to get a feeling for which cutoff 
will be used to generate the paraview (just for the first sample!)

It seems like a distance of 180 for the paraview generation might be 
reasonable.

```{r}
neighbor.dists <- map_dfr(keep.smp$CellID_190411, function(test.fov) {
  
  coords <- cell.data$CellID_190411 %>% filter(fov == test.fov) %>%
    select(c(x, y)) %>% as.data.frame
  
  c(10, 15, 20, 30) %>%
    set_names() %>%
    map_dfc(function(k) {
  dist_k_nn(coords, k)
    }) %>%
  set_names(paste0("dist_to_", c("10", "15", "20", "30")))
  
})

neighbor.dists %>%
  pivot_longer(cols = everything()) %>%
  mutate(name = factor(name, 
                       levels = paste0("dist_to_", c("10", "15", "20", "30")))) %>%
  ggplot() +
  geom_histogram(aes(value), bins = 100) +
  facet_wrap(~ name) +
  lims(x = c(0, 300)) +
  geom_vline(xintercept = 180, col = "blue")
```
