---
title: "EDA MERFISH PREOPTIC"
output:
  workflowr::wflow_html:
    toc: yes
    toc_float: yes
    toc_depth: 2
    code_folding: hide
editor_options: 
  chunk_output_type: inline
bibliography: references.bib
csl: science.csl
---

Setup.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE)
```

```{r}
suppressPackageStartupMessages(library(mistyR))
suppressPackageStartupMessages(library(tidyverse))
```

# Loading and Processing Data

```{r}
data <- read_csv(list.files("data/merfish_preoptic/", full.names = TRUE)[1])
```

# Meta Data

Split meta data and expression data.

```{r}
meta.data = data[,1:9]
meta.data
```

```{r}
expr.data = data[,-c(1:9)] %>%
  select(-starts_with("Blank"))
expr.data
```

Number of samples.

```{r}
samples <-meta.data %>%
  pull(Animal_ID) %>%
  unique()
samples
```

# Statistics per Sample

How many cells per sample?

```{r}
map_dbl(samples, function(smp) {
  meta.data %>%
    filter(Animal_ID == smp) %>%
    nrow()
}) %>%
  tibble(n.cells = ., sample = samples) %>%
  mutate(sample = factor(sample, levels = sort(samples))) %>%
  ggplot() +
  geom_bar(aes(x=n.cells, y=sample), stat="identity") +
  labs(x = "Number of Cells", y = "Sample")
```

```{r}
sample.data <- meta.data %>%
  group_by(Animal_ID) %>%
  summarise(n = n(), Animal_sex = Animal_sex, Behavior = Behavior, .groups = "drop") %>%
  distinct()
sample.data
```

Distribution of cell types per sample

```{r fig.width=12, fig.height=12}
meta.data %>%
  count(Animal_ID, Cell_class) %>%
  ggplot() +
  geom_bar(aes(x = n, y = Cell_class), stat="identity") +
  facet_wrap(~ Animal_ID)
```

Distribution of the markers (difficult to plot since we have 156 markers
and 36 samples!).

So first we will simply plot the mean for each sample for each marker.

```{r fig.width=24, fig.height=6}
expr.data %>%
  mutate(sample = meta.data$Animal_ID) %>%
  pivot_longer(!sample) %>%
  mutate(sample = factor(sample, levels = sort(samples))) %>%
  group_by(sample, name) %>%
  summarise(mean = mean(value), sd = sd(value)) %>%
  mutate(min = mean-sd, max = mean+sd) %>%
  ggplot() +
  geom_point(aes(x = name, y = mean, color = sample)) +
  labs(x = "Markers", y = "Mean Expression") +
  theme(axis.text.x = element_blank())
```

Now let's focus on the spatially variable transcripts.

To that end let's first get the expression and positions per sample.

```{r}
expr.smp <- expr.data %>%
  select(-Fos) %>%
  mutate(sample = meta.data$Animal_ID) %>%
  group_by(sample) %>%
  group_split(.keep=FALSE)

expr.keys <- expr.data %>%
  mutate(sample = meta.data$Animal_ID) %>%
  group_by(sample) %>%
  group_keys()

names(expr.smp) <- expr.keys$sample

pos.smp <- meta.data %>%
  select(Animal_ID, Centroid_X, Centroid_Y) %>%
  group_by(Animal_ID) %>%
  group_split(.keep=FALSE)

pos.keys <- meta.data %>%
  select(Animal_ID, Centroid_X, Centroid_Y) %>%
  group_by(Animal_ID) %>%
  group_keys()

names(pos.smp) <- pos.keys$Animal_ID
```

Quick check.

```{r}
map2_lgl(expr.smp, pos.smp, ~ nrow(.x) == nrow(.y))
```

```{r results='hide'}
spark.results <- map2(expr.smp, pos.smp, ~ SPARK::sparkx(count_in = t(as.matrix(.x)), 
                                                locus_in = as.matrix(.y), 
                                                numCores = 12, 
                                                option = "single", 
                                                verbose = FALSE))

sel.markers <- spark.results %>%
  map(~ .x$res_mtest %>% slice_min(adjustedPval, n = 80) %>% rownames) %>%
  unlist() %>% table %>% as_tibble() %>%
  filter(n > 27) %>%
  rename("marker" = ".") %>%
  pull(marker)
```

```{r}
sel.markers
```

For each sample (animal).

```{r fig.width=24, fig.height=6}
expr.data %>%
  mutate(sample = meta.data$Animal_ID) %>%
  pivot_longer(!sample) %>%
  filter(name %in% sel.markers) %>%
  mutate(sample = factor(sample, levels = sort(samples))) %>%
  ggplot() +
  geom_boxplot(aes(x = name, y = value, color = sample), outlier.shape = NA) +
  lims(y = c(0, 100))
```

For each sex.

```{r fig.width=24, fig.height=6}
expr.data %>%
  mutate(sex = meta.data$Animal_sex) %>%
  pivot_longer(!sex) %>%
  filter(name %in% sel.markers) %>%
  ggplot() +
  geom_boxplot(aes(x = name, y = value, color = sex), outlier.shape = NA) +
  lims(y = c(0, 50))
```

# Spatial Data

Looking at the spatial data (starting with sample = Animal 1)

```{r fig.height=6, fig.width=6}
walk(unique(meta.data$Animal_ID), function(smp) {
  plot <- meta.data %>%
  filter(Animal_ID == smp) %>%
  ggplot() +
  geom_point(aes(x=Centroid_X, y=Centroid_Y, col=Cell_class), size=0.1)
  
  print(plot)
})
```

Now plotting some fewer kinds of cells.

```{r fig.height=6, fig.width=6}
walk(unique(meta.data$Animal_ID), function(smp) {
  plot <- meta.data %>%
  filter(Animal_ID == smp) %>%
  mutate(type = str_extract(Cell_class, "[^0-9]+")) %>%
  ggplot() +
  geom_point(aes(x=Centroid_X, y=Centroid_Y, col=type), size=0.1)
  
  print(plot)
})
```

# Distribution of Distances

Looking at the distances.

Helper function to calculate distances to the neighbors.

```{r}
# helper function for nearest neighbors
dist_k_nn <- function(sample, k) {
  nn <- RANN::nn2(sample, eps=0.0, k = k)
  map_dbl(seq_len(nrow(sample)), function(n) {
    mean(nn$nn.dists[n, ])
  })
}

# helper function for getting direct neighbors
get_neighbors <- function(ddobj, id) {
  dplyr::union(
    ddobj$delsgs$ind1[which(ddobj$delsgs$ind2 == id)],
    ddobj$delsgs$ind2[which(ddobj$delsgs$ind1 == id)]
  )
}
```

First we will look at the distribution of the  mean distances to the direct 
neighbors (as determined by the delauney triangulation) to get a feeling for 
which cutoff value should be used for the generation of the juxtaview. The
corresponding distribution is shown below.
(in blue the cutoff - 30 pixels - which was used in the end).

```{r}
# mean distance to the direct neighbors for each cell
mean.dists.direct.neighbors <- map(pos.smp, function(smp) {
  delaunay <- deldir::deldir(as.data.frame(smp))
  dists <- distances::distances(as.data.frame(smp))
  map_dbl(seq_len(nrow(smp)), function(id) {
    nn <- get_neighbors(delaunay, id)
    mean(dists[nn, id])
  })
}) 

stat.per.sample <- map2_dfr(mean.dists.direct.neighbors,
                            seq_len(length(mean.dists.direct.neighbors)),
                            function(list, smp) {
            data.frame(sample = smp, value = list)
          })

# histogram
ggplot(data.frame(dist = mean.dists.direct.neighbors %>% unlist)) +
  geom_histogram(aes(dist), bins = 100) +
  lims(x = c(0,100)) +
  geom_vline(xintercept = 30, col = "blue")
```

Summary statistics:

```{r}
data.frame(value = mean.dists.direct.neighbors %>% unlist) %>%
  summarise(mean = mean(value), 
            median = median(value), 
            q25 = quantile(value, 0.25),
            q75 = quantile(value, 0.75),
            min = min(value),
            max = max(value))
```

However, a more detailed look across the sample (boxplot) shows that the
cutoff may be too low for certain samples. It may be good to increase this
threshold to 50.

```{r}
# boxplot
stat.per.sample %>%
  mutate(sample = factor(sample)) %>%
  ggplot() +
  geom_boxplot(aes(x=sample, y=value), outlier.shape = NA) +
  geom_hline(yintercept = 30, col="blue") +
  lims(y = c(0, 200))
```

Furthermore we will look at the distribution of the mean distances to the
10 -, 20 -, 40 -, and 80 closest cells to get a feeling
for which cutoff will be used to generate the paraview 
(in blue the cutoff - 120 pixels (46.8 microns) - which was used in the end).

```{r}
neighbor.dists <- map_dfr(pos.smp, function(smp) {
  c(10, 20, 40, 80) %>%
    purrr::set_names() %>%
    map_dfc(function(k) {
  dist_k_nn(smp, k)
    }) %>%
  purrr::set_names(paste0("dist_to_", c("10", "20", "40", "80")))
})

neighbor.dists %>%
  pivot_longer(cols = everything()) %>%
  mutate(name = factor(name, 
                       levels = paste0("dist_to_", c("10", "20", "40", "80")))) %>%
  ggplot() +
  geom_histogram(aes(value), bins = 100) +
  facet_wrap(~ name) +
  lims(x = c(0, 200)) +
  geom_vline(xintercept = 120, col = "blue")
```

Summary statistics:

```{r}
neighbor.dists %>%
  pivot_longer(cols = everything()) %>%
  group_by(name) %>%
  summarise(mean = mean(value), 
            median = median(value), 
            q25 = quantile(value, 0.25),
            q75 = quantile(value, 0.75),
            min = min(value),
            max = max(value))
```


