---
title: "Overview"
site: workflowr::wflow_site
---

This report documents what I have primarily worked on between September 2021
and December 2021. The topic of my internship was "Implement and explore 
different ensemble approaches for training view-specific models for MISTy 
and meta models for late fusion of predictions".

# Tutorial

This Rmarkdown showcases the new implementation of MISTy, where the
function to model each view are passed as an argument to the `run_misty()` 
call. 

```{r eval=FALSE}
misty.views %>% misty.views %>%
  run_misty(model.function = bagged_mars_model, n.bags = 50, degree = 2)
```

The tutorial includes also a section on how to construct those function
(i.e. what is the input and required output).

# Benchmark

This Rmarkdown contains:

a) The results of comparing the performance of different ML algorithms on 
different spatial omics datasets.

b) The results of testing different hyperparameters for certain algorithms
on different spatial omics datasets.

# Frameworks

This Rmarkdown describes why I choose not to use any existing ML framework in
MISTy. In a nutshell, these frameworks are either too slow or too unflexible 
(e.g. not able to return out-of-bag estimates).

# Model API

This Rmarkdown described another way of generalizing the training of 
view-specific models, such that the call looks like this:

```{r eval=FALSE}
misty.views %>% run_misty(method = "bag", learner = "earth", 
                          n.learners = 50, degree = 2)
```


# Random

This Rmarkdown contains results I could not fit into the other categories
thus far.
